# Good practices

```{r, include=FALSE}
WORDS_TO_IGNORE <- "butitsuremakesthingseasiertoread"
# source("knitr-options.R")
# source("spelling-check.R")
```

> “Programs must be written for people to read, and only incidentally for machines to execute.” Harold Abelson

> "Programming is the art of telling another human being what one wants the computer to do." Donald Knuth

> "Let us change our traditional attitude to the construction of programs. Instead of imagining that our main task is to instruct a computer what to do, let us concentrate rather on explaining to human beings what we want a computer to do." Donald Knuth

> "When you write a program, think of it primarily as a work of literature. You're trying to write something that human beings are going to read. Don't think of it primarily as something a computer is going to follow. The more effective you are at making your program readable, the more effective it's going to be: You'll understand it today, you'll understand it next week, and your successors who are going to maintain and modify it will understand it."

## Coding style

Good coding style is like correct punctuation: you can manage without it, butitsuremakesthingseasiertoread. When I answer questions; first, I read the title of the question to see if I can answer the question, secondly, I check the coding style of the question and if the code is too difficult to read, I just move on. Please make your code readable by following e.g. [this coding style](http://style.tidyverse.org/) (examples below come from this guide).

### Comments

In code, use comments to explain the “why” not the “what” or “how”. Each line of a comment should begin with the comment symbol and a single space: `#`.

### Naming

Names are not limited to 8 characters as in some other languages.  Be smart with your naming; be descriptive yet concise. 

Throughout the course will point out some standard naming conventions that are used in R (and other languages).  (Ex. `i` and `j` as row and column indicies)

### Structure

Use commented lines of `-` to create code outline.

### Spacing

Put a space before and after `=` when naming arguments in function calls.
Most infix operators (`==`, `+`, `-`, `<-`, etc.) are also surrounded by
spaces, except those with relatively high precedence: `^`, `:`, `::`, and `:::`. Always put a space after a comma, and never before (just like in regular English).

```{r, eval=FALSE}
# Good
average <- mean((feet / 12) + inches, na.rm = TRUE)
sqrt(x^2 + y^2)
x <- 1:10
base::sum

# Bad
average<-mean(feet/12+inches,na.rm=TRUE)
sqrt(x ^ 2 + y ^ 2)
x <- 1 : 10
base :: sum
```

### Indenting

Curly braces, `{}`, define the the most important hierarchy of R code. To make this hierarchy easy to see, always indent the code inside `{}` by two spaces.

```{r, eval=FALSE}
# Good
if (y < 0 && debug) {
  message("y is negative")
}

if (y == 0) {
  if (x > 0) {
    log(x)
  } else {
    message("x is negative or zero")
  }
} else {
  y ^ x
}

# Bad
if (y < 0 && debug)
message("Y is negative")

if (y == 0)
{
    if (x > 0) {
      log(x)
    } else {
  message("x is negative or zero")
    }
} else { y ^ x }
```

### Long lines

Strive to limit your code to 80 characters per line. This fits comfortably on a 
printed page with a reasonably sized font. If you find yourself running out of 
room, this is a good indication that you should encapsulate some of the work in 
a separate function.

If a function call is too long to fit on a single line, use one line each for 
the function name, each argument, and the closing `)`. 
This makes the code easier to read and to change later. 

```{r, eval=FALSE}
# Good
do_something_very_complicated(
  something = "that",
  requires  = many,
  arguments = "some of which may be long"
)

# Bad
do_something_very_complicated("that", requires, many, arguments,
                              "some of which may be long"
```

### Other

- Use `<-`, not `=`, for assignment. Keep `=` for parameters.

```{r, eval=FALSE}
# Good
x <- 5
system.time(
  x <- rnorm(1e6)
)

# Bad
x = 5
system.time(
  x = rnorm(1e6)
)
```

- Don't put `;` at the end of a line, and don't use `;` to put multiple commands 
on one line.

- Only use `return()` for early returns. Otherwise rely on R to return the result of the last evaluated expression.

```{r, eval=FALSE}
# Good
add_two <- function(x, y) {
  x + y
}

# Bad
add_two <- function(x, y) {
  return(x + y)
}
```

- Use `"`, not `'`, for quoting text. The only exception is when the text already contains double quotes and no single quotes.

```{r, eval=FALSE}
# Good
"Text"
'Text with "quotes"'
'<a href="http://style.tidyverse.org">A link</a>'

# Bad
'Text'
'Text with "double" and \'single\' quotes'
```


## RStudio

Download the latest version of [RStudio](https://www.rstudio.com/products/rstudio/download/#download) (> 1.1) and use it!

Learn more about new features of RStudio v1.1 [there](https://www.rstudio.com/resources/videos/rstudio-1-1-new-features/).

RStudio features:

- everything you can expect from a good IDE
- keyboard shortcuts I use
    1. *Ctrl + Space* (auto-completion, better than *Tab*)
    2. *Ctrl + Up* (command history & search)
    3. *Ctrl + Click* (function source code)
    3. *Ctrl + Enter* (execute line of code)
    4. *Ctrl + Shift + A* (reformat code)
    4. *Ctrl + Shift + C* (comment/uncomment selected lines)
    4. *Ctrl + Shift + K* (knit)
    4. *Ctrl + Shift + B* (build package, website or book)
    4. *Ctrl + Shift + M* (pipe)
    4. *Alt + Shift + K* to see all shortcuts...
- Panels (everything is integrated, including **Git** and a terminal)
- Interactive data importation from files and connections (see [this webinar](https://www.rstudio.com/resources/webinars/importing-data-into-r/))
- Use [code diagnostics](https://support.rstudio.com/hc/en-us/articles/205753617-Code-Diagnostics):

- **R Projects**: 
    - **Meaningful structure** in one folder
    - The working directory automatically switches to the project's folder
    - File tab displays the associated files and folders in the project
    - History of R commands and open files
    - Any settings associated with the project, such as Git settings, are loaded. Note that a *set-up.R* or even a *.Rprofile* file in the project's root directory enable project-specific settings to be loaded each time people work on the project.

<blockquote class="twitter-tweet" data-lang="en" align="center"><p lang="en" dir="ltr">The only two things that make <a href="https://twitter.com/JennyBryan?ref_src=twsrc%5Etfw">\@JennyBryan</a> 😤😠🤯. Instead use projects + here::here() <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> <a href="https://t.co/GwxnHePL4n">pic.twitter.com/GwxnHePL4n</a></p>&mdash; Hadley Wickham (\@hadleywickham) <a href="https://twitter.com/hadleywickham/status/940021008764846080?ref_src=twsrc%5Etfw">December 11 2017</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

Read more at https://www.tidyverse.org/articles/2017/12/workflow-vs-script/ and also see chapter [*Efficient set-up*](https://bookdown.org/csgillespie/efficientR/set-up.html) of book *Efficient R programming*.

## Getting help

### Help yourself, learn how to debug

A basic solution is to print everything, but it usually does not work well on complex problems. A convenient solution to see all the variables' states in your code is to place some `browser()` anywhere you want to check the variables' states.

Learn more with [this book chapter](https://bookdown.org/rdpeng/rprogdatascience/debugging.html), [this other book chapter](http://adv-r.had.co.nz/Exceptions-Debugging.html), [this webinar](https://www.rstudio.com/resources/videos/debugging-techniques-in-rstudio/) and [this RStudio article](https://support.rstudio.com/hc/en-us/articles/205612627-Debugging-with-RStudio).


### External help

Can't remember useful functions? Use [cheat sheets](https://www.rstudio.com/resources/cheatsheets/).

You can search for specific R stuff on https://rseek.org/. You should also read documentations carefully. If you're using a package, search for vignettes and a GitHub repository.

You can also use [Stack Overflow](https://stackoverflow.com/). The most common use of Stack Overflow is when you have an error or a question, you google it, and most of the times the first links are Q/A on Stack Overflow.

You can ask questions on Stack Overflow (using the tag `r`). You need to [make a great R reproducible example](https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example) if you want your question to be answered. Most of the times, while making this reproducible example, you will find the answer to your problem.

If you're confident enough in your R skills, you can go to the next step and [answer questions on Stack Overflow](https://stackoverflow.com/unanswered/tagged/r?tab=newest). It's a good way to increase your skills, or just to [procrastinate while writing a scientific manuscript](https://privefl.github.io/blog/one-month-as-a-procrastinator-on-stack-overflow/).

## Assignement

1. Create an R Project for this class
2. Create the following directories in your project (tip sheet?)(Bonus points if you can do it from R and not RStudio or Windows Explorer)
    * R
    * Rmd
    * figures
    * tables
3. Read Chapters 1-3 of the [Tidyverse Style Guide](http://style.tidyverse.org/index.html)
4. Copy one of your R scripts into your R directory.  (Bonus points if you can do it from R and not RStudio or Windows Explorer)
5. Apply the Style guide to your code.  
6. Apply the "Rule of 3"
    * Create variables as needed
    * Identify code that is used 3 or more times to make functions
    * Identify code that would be useful in 3 or more projects to integrate into a package.
