# Base R Data Structures

## Naming Rules

R has strict rules about what constitutes a valid name. A __syntactic__ name must consist of letters[^letters], digits, `.` and `_`, and can't begin with `_`. Additionally, it can not be one of a list of __reserved words__ like `TRUE`, `NULL`, `if`, and `function` (see the complete list in `?Reserved`). Names that don't follow these rules are called __non-syntactic__ names, and if you try to use them, you'll get an error:

```{r, eval = FALSE}
_abc <- 1
#> Error: unexpected input in "_"

if <- 10
#> Error: unexpected assignment in "if <-"
```

```{block type='rmdwarning'}
While `TRUE` and `FALSE` are reserved words `T` and `F` are not.  However, you can use `T` and `F` as logical.  If someone assigns either of those a different value you will get a __very__ hard to track down bug.  Always spell out the `TRUE` and `FALSE`.
```

[^letters]: Surprisingly, what constitutes a letter is determined by your current locale. That means that the syntax of R code actually differs from computer to computer, and it's possible for a file that works on one computer to not even parse on another!

## Vectors

The most common data structure in R is the vector.  R's vectors can be organised by their dimensionality (1d, 2d, or nd) and whether they're homogeneous or heterogeneous. This gives rise to the five data types most often used in data analysis: 

|    | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Atomic vector | List          |
| 2d | Matrix        | Data frame    |
| nd | Array         |               |

Given an object, the best way to understand what data structures it is composed of is to use `str()`. `str()` is short for structure and it gives a compact, human readable description of any R data structure.

Vectors have three common properties:

* Type, `typeof()`, what it is.
* Length, `length()`, how many elements it contains.
* Attributes, `attributes()`, additional arbitrary metadata.

They differ in the types of their elements: all elements of an atomic vector must be the same type, whereas the elements of a list can have different types.

```{block type='rmdnote'}
`is.vector()` does not test if an object is a vector. Instead it returns TRUE only if the object is a vector with no attributes apart from names. Use `is.atomic(x) || is.list(x)` to test if an object is actually a vector.
```

### Atomic Vectors

There are many "atomic" types of data: `logical`, `integer`, `double` and `character` (in this order, see below). There are also `raw` and `complex` but they are rarely used.

You can't mix types in an atomic vector (you can in a list). Coercion will automatically occur if you mix types:

```{r}
(a <- FALSE)
typeof(a)

(b <- 1:10)
typeof(b)
c(a, b)         ## FALSE is coerced to integer 0

(c <- 10.5)
typeof(c)
(d <- c(b, c))  ## coerced to double

c(d, "a")       ## coerced to character

c(list(1), "a")

50 < "7"
```

You can force coercion with `as.logical`, `as.integer`, `as.double`, `as.numeric`, and `as.character`. Most of the time the coercion rules are straight forward, but not always.

```{r}
x <- c(TRUE, FALSE)
typeof(x)

as.integer(x)
as.numeric(x)
as.character(x)
```

However, coercion is not associative.

```{r}
x <- c(TRUE, FALSE)

x2 <- as.integer(x)
x3 <- as.numeric(x2)
as.character(x3)
```

What would you expect this to return?

```{r, eval = FALSE}
x <- c(TRUE, FALSE)

as.integer(as.character(x))
```

You can test for an "atomic" types of data with: `is.logical`, `is.integer`, `is.double`, `is.numeric`[^numeric], and `is.character`.

[^numeric]: `is.numeric()` is a general test for the “numberliness” of a vector and returns TRUE for both integer and double vectors. It is not a specific test for double vectors, which are often called numeric.

```{r}
x <- c(TRUE, FALSE)

is.logical(x)
is.integer(x)
```

What would you expect these to return?

```{r, eval = FALSE}
x <- 2

is.integer(x)
is.numeric(x)
is.double(x)
```

Missing values are specified with `NA`, which is a logical vector of length 1. `NA` will always be coerced to the correct type if used inside `c()`, or you can create `NA`s of a specific type with `NA_real_` (a double vector), `NA_integer_` and `NA_character_`.

### Lists

Lists are different from atomic vectors because their elements can be of any type, including other lists. Lists can contain complex objects so it's not possible to pick one visual style that works for every list.  You construct lists by using `list()` instead of `c()`: 

```{r}
x <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
str(x)
```

Lists are sometimes called __recursive__ vectors, because a list can contain other lists. This makes them fundamentally different from atomic vectors.

```{r}
x <- list(list(list(list(1))))
str(x)
is.recursive(x)
```

`c()` will combine several lists into one. If given a combination of atomic vectors and lists, `c()` will coerce the vectors to lists before combining them. Compare the results of `list()` and `c()`:

```{r}
x <- list(list(1, 2), c(3, 4))
y <- c(list(1, 2), c(3, 4))
str(x)
str(y)
```

The `typeof()` a list is `list`. You can test for a list with `is.list()` and coerce to a list with `as.list()`. You can turn a list into an atomic vector with `unlist()`. If the elements of a list have different types, `unlist()` uses the same coercion rules as `c()`.

Lists are used to build up many of the more complicated data structures in R. For example, both data frames (described in data frames) and linear models objects (as produced by `lm()`) are lists

### `NULL`

Closely related to vectors is `NULL`, a singleton object often used to represent a vector of length 0. `NULL` is different than `NA`.  For a good explanation of the differences see [this](https://www.r-bloggers.com/r-na-vs-null/) blog post.

### Attributes

All objects can have arbitrary additional attributes, used to store metadata about the object. Attributes can be thought of as a named list[^pairlist] (with unique names). Attributes can be accessed individually with `attr()` or all at once (as a list) with `attributes()`. 

[^pairlist]: The reality is a little more complicated: attributes are actually stored in something called pairlists, which can you learn more about in [Advanced R](http://adv-r.had.co.nz)


```{r}
a <- 1:3
attr(a, "x") <- "abcdef"
attr(a, "y") <- 4:6
attr(a, "z") <- list(list())
str(attributes(a))
```

The `structure()` function returns a new object with modified attributes. Care must be taken with attributes since, by default, most attributes are lost when modifying a vector.

```{r}
attributes(a[1])
attributes(sum(a))
```

The only attributes not lost are the three most important:

* Names, a character vector giving each element a name. 

* Dimensions, used to turn vectors into matrices and arrays.

* Class, used to implement the S3 object system.
 
Each of these attributes has a specific accessor function to get and set values. When working with these attributes, use `names(x)`, `dim(x)`, and `class(x)`, not `attr(x, "names")`, `attr(x, "dim")`, and `attr(x, "class")`.

#### Names

You can name a vector in a couple[^named-vectors] ways:

* When creating it: `x <- c(a = 1, b = 2, c = 3)`.

* By modifying an existing vector in place: 
  `x <- 1:3; names(x) <- c("a", "b", "c")`.
  
Named vectors are a great way to make an easy, human readable look up table.  We will see this use case extensively when we get to data visualizations.  

[^named-vectors]: There are a couple less common ways. See [Advanced R](http://adv-r.had.co.nz)

### Factors

One important use of attributes is to define factors. A factor is a vector that can contains only predefined values, and is used to store categorical data. Factors are built on top of **integer vectors** using two attributes: the `class`, "factor", which makes them behave differently from regular integer vectors, and the `levels`, which defines the set of allowed values.  Factors can also have labels which effect how the factors are displayed.  By default the labels are the same as the levels.

The order of the levels of a factor can be set using the levels argument to `factor()`. This can be important in linear modelling because the first level is used as the baseline level. This feature can also be used to customize order in plots that include factors, since by default factors are plotted in the order of their levels. Labels are also useful in plotting where you want the displayed text to be different than the underlying representation.

Factors are useful when you know the possible values a variable may take, even if you don't see all values in a given data set. Using a factor instead of a character vector makes it obvious when some groups contain no observations:

```{r}
gender_char <- c("m", "m", "m")
gender_factor <- factor(gender_char, levels = c("m", "f"))

gender_char
table(gender_char)
gender_factor
table(gender_factor)
# See the underlying representation of a factor
unclass(gender_factor)

gender_factor2 <- factor(gender_char, levels = c("m", "f"), labels = c("Male", "Female"))
gender_factor2
table(gender_factor2)
# See the underlying representation of a factor
unclass(gender_factor2)
```

While factors look like (and often behave like) character vectors, they are actually **integers**. Be careful when treating them like strings. Some string methods (like `gsub()` and `grepl()`) will coerce factors to strings, while others (like `nchar()`) will throw an error, and still others (like `c()`) will use the underlying integer values. For this reason, it is best to explicitly convert factors to character vectors if you need string-like behavior.  

Unfortunately, many base R functions (like `read.csv()` and `data.frame()`) automatically convert character vectors to factors. This is sub-optimal, because there's no way for those functions to know the set of all possible levels or their optimal order. Instead, use the argument `stringsAsFactors = FALSE` to suppress this behavior, and then manually convert character vectors to factors using your knowledge of the data only when you need the behavior of factors.  

Factors tend to be most useful in data visualization and table creations where you want to report all categories but some categories may not be present in your data, or when you want to order the categories in something other than the default ordering.  We will revisit factors and there usefulness later when we study the tidyverse and in particular the forcats package.

### Matrices and arrays

Adding a `dim` attribute to an atomic vector allows it to behave like a multi-dimensional __array__. A special case of the array is the __matrix__, which has two dimensions. Matrices are used commonly as part of the mathematical machinery of statistics. Arrays are much rarer, but worth being aware of.

Matrices and arrays are created with `matrix()` and `array()`, or by using the assignment form of `dim()`:

```{r}
# Two scalar arguments to specify rows and columns
a <- matrix(1:12, ncol = 3, nrow = 4)
a
# One vector argument to describe all dimensions
b <- array(1:12, c(2, 3, 2))
b

# You can also modify an object in place by setting dim()
vec <- 1:12
vec
class(vec)

dim(vec) <- c(3, 4)
vec
class(vec)

dim(vec) <- c(3, 2, 2)
vec
class(vec)
```

`length()` and `names()` have high-dimensional generalizations:

* `length()` generalizes to `nrow()` and `ncol()` for matrices, and `dim()`
  for arrays.

* `names()` generalizes to `rownames()` and `colnames()` for matrices, and
  `dimnames()`, a list of character vectors, for arrays. 

`c()` generalizes to `cbind()` and `rbind()` for matrices, and to `abind::abind()` for arrays. You can transpose a matrix with `t()`; the generalized equivalent for arrays is `aperm()`.

You can test if an object is a matrix or array using `is.matrix()` and `is.array()`, or by looking at the length of the `dim()`. `as.matrix()` and `as.array()` make it easy to turn an existing vector into a matrix or array.

Vectors are not the only 1-dimensional data structure. You can have matrices with a single row or single column, or arrays with a single dimension. They may print similarly, but will behave differently. The differences aren't too important, but it's useful to know they exist in case you get strange output from a function (`tapply()` is a frequent offender). As always, use `str()` to reveal the differences.

Matrices and arrays are most useful for mathematical calculations (particularly when fitting models); lists and data frames are a better fit for most other programming tasks in R.

### Data Frames

A data frame is the most common way of storing data in R, and if used systematically makes data analysis easier. Under the hood, a data frame is a list of equal-length vectors. This makes it a 2-dimensional structure, so it shares properties of both the matrix and the list.  This means that a data frame has `names()`, `colnames()`, and `rownames()`, although `names()` and `colnames()` are the same thing. The `length()` of a data frame is the length of the underlying list and so is the same as `ncol()`; `nrow()` gives the number of rows. 
You can subset a data frame like a 1d structure (where it behaves like a list), or a 2d structure (where it behaves like a matrix), we will discuss this further when we discuss subsetting.

#### Creation

You create a data frame using `data.frame()`, which takes named vectors as input:

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
str(df)
```

```{block2, type='rmdwarning'}
Beware data.frame()'s default behavior which turns strings into factors. Use `stringsAsFactors = FALSE` to suppress this behavior.
```

```{r}
df <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"),
  stringsAsFactors = FALSE)
str(df)
```

#### Testing and coercion

Because a `data.frame` is an S3 class, its type reflects the underlying vector used to build it: the list. To check if an object is a data frame, use `is.data.frame()`:

```{r}
is.data.frame(df)
```

You can coerce an object to a data frame with `as.data.frame()`:

* A vector will create a one-column data frame.

* A list will create one column for each element; it's an error if they're 
  not all the same length.
  
* A matrix will create a data frame with the same number of columns and 
  rows as the matrix.
  
The automatic coercion that causes the most problems is if you select a single column of a data.frame. R will coerce the column to an atomic vector, which generally is not what you want[^data.frame].

```{r}
(x1 <- df[, "x"])
str(x1)

(x2 <- df[, "y", drop = FALSE])
str(x2)

```

[^data.frame]: We'll revisit this when we get into R for Data Science and discuss tibbles


#### Combining data frames

You can combine data frames using `cbind()` and `rbind()`: 

```{r}
cbind(df, data.frame(z = 3:1))
rbind(df, data.frame(x = 10, y = "z"))
```

When combining column-wise, the number of rows must match, but row names are ignored. When combining row-wise, both the number and names of columns must match.

It's a common mistake to try and create a data frame by `cbind()`ing vectors together. This is unlikely to do what you want because `cbind()` will create a matrix unless one of the arguments is already a data frame. Instead use `data.frame()` directly:

```{r}
# This is always a mistake
bad <- data.frame(cbind(a = 1:2, b = c("a", "b")))
str(bad)

good <- data.frame(a = 1:2, b = c("a", "b"))
str(good)
```

#### List and matrix columns

Since a data frame is a list of vectors, it is possible for a data frame to have a column that is a list. This is a powerful technique because a list can contain any other R object. This means that you can have a column of data frames, or model objects, or even functions!  We will see this again when we discuss tidy data.

```{r}
df <- data.frame(x = 1:3)
df$y <- list(1:2, 1:3, 1:4)
df
```

However, when a list is given to `data.frame()`, it tries to put each item of the list into its own column, so this fails:

```{r, error = TRUE}
data.frame(x = 1:3, y = list(1:2, 1:3, 1:4))
```

A workaround is to use `I()`, which causes `data.frame()` to treat the list as one unit:

```{r}
dfl <- data.frame(x = 1:3, y = I(list(1:2, 1:3, 1:4)))
str(dfl)
```

`I()` adds the `AsIs` class to its input, but this can usually be safely ignored. 

Similarly, it's also possible to have a column of a data frame that's a matrix or array, as long as the number of rows matches the data frame:

```{r}
dfm <- data.frame(x = 1:3 * 10, y = I(matrix(1:9, nrow = 3)))
str(dfm)
```

Use list and array columns with caution. Many functions that work with data frames assume that all columns are atomic vectors, and the printed display can be confusing.

```{r}
dfl[2, ]
dfm[2, ]
```
