# Iteration

## Control Structures

Control structures allow you to put some "logic" into your R code, rather than just always executing the same R code every time. Control structures allow you to respond to inputs or to features of the data and execute different R expressions accordingly.

Commonly used control structures are

- `if` and `else`: testing a condition and acting on it

- `for`: execute a loop a fixed number of times 

- `while`: execute a loop _while_ a condition is true 

- `repeat`: execute an infinite loop (must `break` out of it to stop)

- `break`: break the execution of a loop

- `next`: skip an iteration of a loop

### `if`-`else`

The `if`-`else` combination is probably the most commonly used control
structure in R (or perhaps any language). This structure allows you to
test a condition and act on it depending on whether it's true or
false. 

For starters, you can just use the `if` statement.

```r
if(<condition>) {
        # do something
} 
# Continue with rest of code
```

The above code does nothing if the condition is false. If you have an
action you want to execute when the condition is false, then you need
an `else` clause.

```r
if(<condition>) {
        # do something
} 
else {
        # do something else
}
```

You can have a series of tests by following the initial `if` with any
number of `else if`s.

```r
if(<condition1>) {
        # do something
} else if(<condition2>)  {
        # do something different
} else {
        # do something else different
}
```

```{block type='rmdwarning'}
There is also an `ifelse` function which is vectorized version.  It is essentially an `if`-`else` wrapped in a `for` loop so that the condition, and action, is performed on each element in a vector.
```


### `for` Loops

For loops are pretty much the only looping construct that you will
need in R. While you may occasionally find a need for other types of
loops, in my experience doing data analysis, I've found very few
situations where a for loop wasn't sufficient. 

In R, for loops take an iterator variable and assign it successive
values from a sequence or vector. For loops are most commonly used for
iterating over the elements of an object (list, vector, etc.)

The following three loops all have the similar behavior.

```{r}
x <- c("a", "b", "c", "d")

for(i in 1:length(x)) {
        ## Print out each element of 'x'
        print(x[i])  
}
```

The `seq_along()` function is commonly used in conjunction with for
loops in order to generate an integer sequence based on the length of
an object (in this case, the object `x`).

```{r}
## Generate a sequence based on length of 'x'
for(i in seq_along(x)) {   
        print(x[i])
}
```

It is not necessary to use an index-type variable. 

```{r}
for(letter in x) {
        print(letter)
}
```

```{block2, type='rmdtip'}
Nested loops are commonly needed for multidimensional or hierarchical
data structures (e.g. matrices, lists). Be careful with nesting
though. Nesting beyond 2 to 3 levels often makes it difficult to
read/understand the code. If you find yourself in need of a large
number of nested loops, you may want to break up the loops by using
functions (discussed later).
```

### `while` Loops

While loops begin by testing a condition. If it is true, then they
execute the loop body. Once the loop body is executed, the condition
is tested again, and so forth, until the condition is false, after
which the loop exits.

```{r}
count <- 0
while(count < 10) {
        print(count)
        count <- count + 1
}
```

```{block2, type='rmdwarning'}
While loops can potentially result in infinite loops if not written
properly. Use with care!
```

Sometimes there will be more than one condition in the test.

```{r}
z <- 5
set.seed(1)

while(z >= 3 && z <= 10) {
        coin <- rbinom(1, 1, 0.5)
        
        if(coin == 1) {  ## random walk
                z <- z + 1
        } else {
                z <- z - 1
        } 
}
print(z)
```

Conditions are always evaluated from left to right. For example, in
the above code, if `z` were less than 3, the second test would not
have been evaluated.

### `repeat` Loops

`repeat` initiates an infinite loop right from the start. These are
not commonly used in statistical or data analysis applications but
they do have their uses. The only way to exit a `repeat` loop is to
call `break`.

One possible paradigm might be in an iterative algorithm where you may
be searching for a solution and you don't want to stop until you're
close enough to the solution. In this kind of situation, you often
don't know in advance how many iterations it's going to take to get
"close enough" to the solution.

```{r,eval=FALSE,prompt=FALSE}
x0 <- 1
tol <- 1e-8

repeat {
        x1 <- computeEstimate()
        
        if(abs(x1 - x0) < tol) {  ## Close enough?
                break
        } else {
                x0 <- x1
        } 
}
```

```{block type='rmdnote'}
The above code will not run since the `computeEstimate()` function is not defined. I just made it up for the purposes of this demonstration.
```

The loop above is a bit dangerous because there's no guarantee it will ever stop. You could get in a situation where the values of `x0` and `x1`
oscillate back and forth and never converge. Better to set a hard
limit on the number of iterations by using a `for` loop and then
report whether convergence was achieved or not.

### `next`, `break`

While not used very often it's nice to know about these.

`next` is used to skip an iteration of a loop. 

```{r,eval=FALSE,prompt=FALSE}
for(i in 1:100) {
        if(i <= 20) {
                ## Skip the first 20 iterations
                next                 
        }
        ## Do something here
}
```

`break` is used to exit a loop immediately, regardless of what
iteration the loop may be on.

```{r,eval=FALSE,prompt=FALSE}
for(i in 1:100) {
      print(i)

      if(i > 20) {
              ## Stop loop after 20 iterations
              break  
      }		
}
```

### Looping 

For loops are so common that that R has some functions which implement looping in a compact form to make your life easier.  For a more in depth look see [this](https://bookdown.org/rdpeng/rprogdatascience/loop-functions.html)

*  `apply` is generic: applies a function to a matrix's rows or columns (or, more generally, to dimensions of an array)
*  `lapply` is a list apply which acts on a list or vector and returns a list.
*  `sapply` is a simple lapply but defaults to returning a vector (or matrix) if possible.
*  `vapply` is a verified apply.  This is a sapply with the return object type pre-specified.
*  `rapply` is a recursive apply for nested lists, i.e. lists within lists
*  `tapply` is a tagged apply where the tags identify the subsets to apply a function
*  `mapply` is a multivariate apply for functions that have multiple arguments.
*  `Map` is a wrapper to mapply with SIMPLIFY = FALSE, so it is guaranteed to return a list.
*  `replicate` is a wrapper around sapply for repeated evaluation of an expression

```{r}
# Two dimensional matrix
M <- matrix(sample(1:16), 4, 4)
M
# apply min to rows
apply(M, 1, min)
# apply max to columns
apply(M, 2, max)
```

If you want row/column means or sums for a 2D matrix, be sure to investigate the highly optimized, lightning-quick `colMeans`, `rowMeans`, `colSums`, `rowSums`.

```{r}
x <- list(a = 1, b = 1:3, c = 10:25)
x
lapply(x, FUN = length) 
sapply(x, FUN = length) 
vapply(x, FUN = length, FUN.VALUE = 0L) 

x <- 1:20
y <- factor(rep(letters[1:5], each = 4)) # a vector of the same length as x
tapply(x, y, sum) 

# Sums the 1st elements, the 2nd elements, etc. 
mapply(sum, 1:5, 1:5, 1:5) 

# find the mean of 10 random normal variables, 5 times
replicate(5, mean(rnorm(10)))
```

`tapply` is in a similar spirit to a common data analysis paradigm called split-apply-combine where we split our data set based on a group, apply a function or code to it, and combine the results back together.  We will revisit this paradigm in greater detail when we get to _R for Data Science_.  


