# (PART) Base R Basics {-}

# R Basics {#baser-rbasics}

Here is  a quick overview of the basics.  Next we'll dive deep into R's basic data structures and then how to subset these data structures.  This will give us a good overview of base R and the background needed to dive into __R for Data Science__.

The three most important functions in R `?`, `??`, and `str`:

- `?topic` provides access to the documentation for _topic_.
- `??topic` searches the documentation for _topic_.
- `str` displays the structure of an R object in human readable form.

See this [vocabulary list](http://adv-r.had.co.nz/Vocabulary.html) for a good starting point on the basics functions in base R and some important libraries.  

A book to learn the basics is [R Programmig for Data Science](https://bookdown.org/rdpeng/rprogdatascience/)

In R there three basic constructs; objects, functions, and environments:

## Assignment Operators

We saw this is Coding Style. Use `<-` for assignment and use `=` for parameters.  While you can use `=` for assignment it is generally considered bad practice.

## Objects

### Vector

You create a vector with `c`.

```{r}
v <- c("my", "first", "vector")
v

# length of our vector
length(v)
```

There are several shortcut functions for common vector creation.
```{r}
# create an ordered sequence
2:10
9:3

# generate regular sequences
seq(1, 20, by = 3)

# replicate a number n times
rep(3, times = 4)

# arguments are generally vectorized
rep(1:3, times = 3:1)

# common mistake using 1:length(n) in loops
# but if n = 0
1:0

# use seq_len(n) instead and the loop won't execute
seq_len(0)

# another common mistake
n <- 6
1:n+1        # is (1:n) + 1, so 2:(n + 1)
1:(n+1)      # usually what is meant
seq_len(n+1) # a better way
```

### Matrix

Matrices are 2D vectors, with all elements of the same type.  Generally used for mathematics.

```{r}
# fill in column order (default)
matrix(1:12, nrow = 3)

# fill in row order
matrix(1:12, nrow = 3, byrow = TRUE)

# can also specify the number of columns instead
matrix(1:12, ncol = 3)
```

You find the dimensions of a matrix with `nrow`, `ncol`, and `dim`

```{r}
m <- matrix(1:12, ncol = 3)
dim(m)
nrow(m)
ncol(m)
```

### List

A list is a generic vector containing other objects.  These do __NOT__ have to be the same type or the same length.

```{r}
s <- c("aa", "bb", "cc", "dd", "ee") 
b <- c(TRUE, FALSE, TRUE, FALSE, FALSE) 
# x contains copies of n, s, b and our matrix from above
x <- list(n = c(2, 3, 5) , s, b, 3, m)   
x

# length gives you length of the list not the elements in the list
length(x)
```

We'll discuss lists in detail in the next chapter.

### Data frame

A data frame is a list with each vector of the same length.  This is the main data structure used and is analogous to a data set in SAS.  While these __look__ like matrices they behave very different.

```{r}
df = data.frame(n = c(2, 3, 5),
                s = c("aa", "bb", "cc"),
                b = c(TRUE, FALSE, TRUE),
                y = v
                )       # df is a data frame 
df

# dimensions
dim(df)
nrow(df)
ncol(df)
length(df)
```

We'll discuss data frames in greater detail in the next chapter.

## Comparision 

```{r table-logicalOps, tidy=FALSE, echo=FALSE}

log_ops <- data.frame(Operator = c(">", ">=", "<", "<=", "==", "!="),
                      Description = c("greater than",
                                      "greater than or equal to",
                                      "less than",
                                      "less than or equal to",
                                      "exactly equal to",
                                      "not equal to "))

knitr::kable(
  log_ops, booktabs = TRUE,
  caption = 'Logical Operators'
)
```

```{r}
v <- 1:12
v[v > 9]
```

Equality can be tricky to test for since real numbers can't be expressed exactly in computers.

```{r}
x <- sqrt(2)
(y <- x^2)
y == 2
print(y, digits = 20)
all.equal(y, 2)          ## equality with some tolerance
all.equal(y, 3)
isTRUE(all.equal(y, 3))  ## if you want a boolean, use isTRUE()
```

## Logical and sets

```{r}
x <- c(TRUE, FALSE)
df <- data.frame(expand.grid(x, x))
names(df) <- c("x", "y")
df$and  <- df$x & df$y     # logical and
df$or   <- df$x | df$y     # logical or
df$notx <- !df$x           # negation
df$xor  <- xor(df$x, df$y) # exlusive or
df
```

R has two versions of the logical operators `&` and `&&`  (`|` and `||`).  The single version is the vectorized version while the the double version returns a length-one vector.  Use the double version in logical control structures (if, for, while, etc).  
```{r}
# TRUE/FALSE and each element
TRUE & c(TRUE, FALSE)
FALSE & c(TRUE, FALSE)
# TRUE/FALSE and first element
TRUE && c(TRUE, FALSE)
FALSE && c(TRUE, FALSE)
# TRUE/FALSE or each element
TRUE | c(TRUE, FALSE)
FALSE | c(TRUE, FALSE)
# TRUE/FALSE or first element
TRUE || c(TRUE, FALSE)
FALSE || c(TRUE, FALSE)
```


This is a common source of bugs in control structures (if, for, while, etc) where you must have a single TRUE / FALSE. 

```{block type='rmdcaution'}
`=` is used for assignment while `==` is used for comparison. A common bug is to use `=` instead of `==` inside a control structure. 
```


It also has useful helpers `any` and `all`

```{r}
x <- c(FALSE, FALSE, FALSE, TRUE)
any(x)
all(x)
all(!x[1:3])
```

And also some useful __set__ operations `intersect`, `union`, `setdiff`, `setequal`

```{r}
x <- 1:5
y <- 3:7

intersect(x, y) # in x and in y
union(x, y)     # different than c()
c(x,y)          # not a set operation
setdiff(x, y)   # in x but not in y
setdiff(y, x)   # in y but not in x
setequal(x, y)
z <- 5:1
setequal(x, z)
```

## Control Structures

Control structures allow you to put some "logic" into your R code, rather than just always executing the same R code every time. Control structures allow you to respond to inputs or to features of the data and execute different R expressions accordingly.

Commonly used control structures are

- `if` and `else`: testing a condition and acting on it

- `for`: execute a loop a fixed number of times 

- `while`: execute a loop _while_ a condition is true 

- `repeat`: execute an infinite loop (must `break` out of it to stop)

- `break`: break the execution of a loop

- `next`: skip an iteration of a loop

### `if`-`else`

The `if`-`else` combination is probably the most commonly used control
structure in R (or perhaps any language). This structure allows you to
test a condition and act on it depending on whether it's true or
false. 

For starters, you can just use the `if` statement.

```r
if(<condition>) {
        # do something
} 
# Continue with rest of code
```

The above code does nothing if the condition is false. If you have an
action you want to execute when the condition is false, then you need
an `else` clause.

```r
if(<condition>) {
        # do something
} 
else {
        # do something else
}
```

You can have a series of tests by following the initial `if` with any
number of `else if`s.

```r
if(<condition1>) {
        # do something
} else if(<condition2>)  {
        # do something different
} else {
        # do something else different
}
```
### `for` Loops

For loops are pretty much the only looping construct that you will
need in R. While you may occasionally find a need for other types of
loops, in my experience doing data analysis, I've found very few
situations where a for loop wasn't sufficient. 

In R, for loops take an iterator variable and assign it successive
values from a sequence or vector. For loops are most commonly used for
iterating over the elements of an object (list, vector, etc.)

The following three loops all have the similar behavior.

```{r}
x <- c("a", "b", "c", "d")

for(i in 1:length(x)) {
        ## Print out each element of 'x'
        print(x[i])  
}
```

The `seq_along()` function is commonly used in conjunction with for
loops in order to generate an integer sequence based on the length of
an object (in this case, the object `x`).

```{r}
## Generate a sequence based on length of 'x'
for(i in seq_along(x)) {   
        print(x[i])
}
```

It is not necessary to use an index-type variable. 

```{r}
for(letter in x) {
        print(letter)
}
```

```{block2, type='rmdtip'}
Nested loops are commonly needed for multidimensional or hierarchical
data structures (e.g. matrices, lists). Be careful with nesting
though. Nesting beyond 2 to 3 levels often makes it difficult to
read/understand the code. If you find yourself in need of a large
number of nested loops, you may want to break up the loops by using
functions (discussed later).
```

### `while` Loops

While loops begin by testing a condition. If it is true, then they
execute the loop body. Once the loop body is executed, the condition
is tested again, and so forth, until the condition is false, after
which the loop exits.

```{r}
count <- 0
while(count < 10) {
        print(count)
        count <- count + 1
}
```

```{block2, type='rmdwarning'}
While loops can potentially result in infinite loops if not written
properly. Use with care!
```

Sometimes there will be more than one condition in the test.

```{r}
z <- 5
set.seed(1)

while(z >= 3 && z <= 10) {
        coin <- rbinom(1, 1, 0.5)
        
        if(coin == 1) {  ## random walk
                z <- z + 1
        } else {
                z <- z - 1
        } 
}
print(z)
```

Conditions are always evaluated from left to right. For example, in
the above code, if `z` were less than 3, the second test would not
have been evaluated.

### `repeat` Loops

`repeat` initiates an infinite loop right from the start. These are
not commonly used in statistical or data analysis applications but
they do have their uses. The only way to exit a `repeat` loop is to
call `break`.

One possible paradigm might be in an iterative algorithm where you may
be searching for a solution and you don't want to stop until you're
close enough to the solution. In this kind of situation, you often
don't know in advance how many iterations it's going to take to get
"close enough" to the solution.

```{r,eval=FALSE,prompt=FALSE}
x0 <- 1
tol <- 1e-8

repeat {
        x1 <- computeEstimate()
        
        if(abs(x1 - x0) < tol) {  ## Close enough?
                break
        } else {
                x0 <- x1
        } 
}
```

```{block type='rmdnote'}
The above code will not run since the `computeEstimate()` function is not defined. I just made it up for the purposes of this demonstration.
```

The loop above is a bit dangerous because there's no guarantee it will ever stop. You could get in a situation where the values of `x0` and `x1`
oscillate back and forth and never converge. Better to set a hard
limit on the number of iterations by using a `for` loop and then
report whether convergence was achieved or not.

### `next`, `break`

While not used very often it's nice to know about these.

`next` is used to skip an iteration of a loop. 

```{r,eval=FALSE,prompt=FALSE}
for(i in 1:100) {
        if(i <= 20) {
                ## Skip the first 20 iterations
                next                 
        }
        ## Do something here
}
```

`break` is used to exit a loop immediately, regardless of what
iteration the loop may be on.

```{r,eval=FALSE,prompt=FALSE}
for(i in 1:100) {
      print(i)

      if(i > 20) {
              ## Stop loop after 20 iterations
              break  
      }		
}
```

### Looping 

For loops are so common that that R has some functions which implement looping in a compact form to make your life easier.  For a more in depth look see [this](https://bookdown.org/rdpeng/rprogdatascience/loop-functions.html)

*  `apply` is generic: applies a function to a matrix's rows or columns (or, more generally, to dimensions of an array)
*  `lapply` is a list apply which acts on a list or vector and returns a list.
*  `sapply` is a simple lapply but defaults to returning a vector (or matrix) if possible.
*  `vapply` is a verified apply.  This is a sapply with the return object type prespecified.
*  `rapply` is a recursive apply for nested lists, i.e. lists within lists
*  `tapply` is a tagged apply where the tags identify the subsets to apply a function
*  `mapply` is a multivariate apply for functions that have multiple arguments.
*  `Map` is a wrapper to mapply with SIMPLIFY = FALSE, so it is guaranteed to return a list.
*  `replicate` is a wrapper around sapply for repeated evaluation of an expression

```{r}
# Two dimensional matrix
M <- matrix(sample(1:16), 4, 4)
M
# apply min to rows
apply(M, 1, min)
# apply max to columns
apply(M, 2, max)
```

If you want row/column means or sums for a 2D matrix, be sure to investigate the highly optimized, lightning-quick `colMeans`, `rowMeans`, `colSums`, `rowSums`.

```{r}
x <- list(a = 1, b = 1:3, c = 10:25)
x
lapply(x, FUN = length) 
sapply(x, FUN = length) 
vapply(x, FUN = length, FUN.VALUE = 0L) 

x <- 1:20
y <- factor(rep(letters[1:5], each = 4)) # a vector of the same length as x
tapply(x, y, sum) 

# Sums the 1st elements, the 2nd elements, etc. 
mapply(sum, 1:5, 1:5, 1:5) 

# find the mean of 10 random normal variables, 5 times
replicate(5, mean(rnorm(10)))
```

`tapply` is in a simalar spirit to a common data analysis paradigm called split-apply-combine where we split our data set based on a group, apply a function or code to it, and combine the results back together.  We will revisit this paradigm in greater detail when we get to _R for Data Science_.  


## Vectorization & Recycling

Many operations in R are _vectorized_, meaning that operations occur in parallel in certain R objects. This allows you to write code that is efficient, concise, and easier to read than in non-vectorized languages.

The simplest example is when adding two vectors together.

```{r}
x <- 1:3
y <- 11:13
z <- x + y
z
```

In most other languages you would have to do something like
```{r}
z <- numeric(length(x))

for(i in seq_along(x)) {
      z[i] <- x[i] + y[i]
}
z
```

We saw a a form of vectorization above in the logical operators.

```{r}
x
x > 2
x[x > 2]
```

Matrix operations are also vectorized, making for nice compact notation. This way, we can do element-by-element operations on matrices without having to loop over every element.

```{r}
x <- matrix(1:4, 2, 2)
y <- matrix(rep(10, 4), 2, 2)
x
y
x * y  # element-wise multiplication
x / y  # element-wise division
x %*% y  # true matrix multiplication
```

R also recycles arguments.  

```{r}
x <- 1:10
z <- x + .1  # add .1 to each element
z
```

While you usually either want the same length vector or a length one vector.  You are not limited to just these options.

```{r}
x <- 1:10
y <- x + c(.1, .2) 
y
z <- x + c(.1, .2, .3)
z
```

### Example

One (not so good) way to estimate `pi` is through Monte-Carlo simulation.

Suppose we wish to estimate the value of `pi` using a Monte-Carlo method.  Essentially, we throw darts at the unit square and count the number of darts that fall within the unit circle.  We'll only deal with quadrant one.  Thus the $Area = \frac{\pi}{4}$

Monte-Carlo pseudo code:

1. Initialize `hits = 0`
2. __for i in 1:N__
3.    Generate two random numbers, $U_1$ and $U_2$, between 0 and 1
4.    If $U_1^2 + U_2^2 < 1$, then `hits = hits + 1`
5. __end for__
6. Area estimate = `hits / N`
7. $\hat{pi} = 4 * Area Estimate$

```{r cache=TRUE}
pi_naive <- function(N) {
  hits <- 0
  for(i in seq_len(N)) {
    U1 <- runif(1)
    U2 <- runif(1)
    if ((U1^2 + U2^2) < 1) {
      hits <- hits + 1
    }
  }
  
  4*hits/N
}
N <- 1e6
system.time(pi_naive(N))
```

That's a long run time (and bad estimate).  Let's vectorize it.

```{r}
pi_vect <- function(N) {
  U1 <- runif(N)
  U2 <- runif(N)
  hits <- sum(U1^2 + U2^2 < 1)
  4*hits/N
}
system.time(pi_vect(N))
```

That is ~20x speed up.

## Function Basics

> To understand computations in R, two slogans are helpful:  
> - Everything that exists is an object.  
> - Everything that happens is a function call.  
> 
> -- John Chambers

Functions are an central part of robust R programming and we will spend a significant amount of time writing functions.

Functions in R are "first class objects", which means that they can be
treated much like any other R object. Importantly,

- Functions can be passed as arguments to other functions. This is very handy for the various apply functions, like `lapply()` and `sapply()`.

- Functions can be nested, so that you can define a function inside of
  another function

If you're familiar with common language like C, these features might appear a bit strange. However, they are really important in R and can be useful for data analysis.

- Functions are a means of **abstraction**. A concept/computation is
  encapsulated/isolated from the rest with a function.
- Functions should **do one thing**, and do it well (compute, or plot,
  or save, ... not all in one go).
- **Side effects**: your functions should not have any (unless, of
  course, that is the main point of that function - plotting, write to
  disk, ...). Functions shouldn't make any changes in any
  environment. The only return their output.
- **Do not use global variables**. Everything the function needs is
  being passed as an argument. Function must be **self-contained**.
- Function streamline code and process

Advice from the [R Inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf):

Make your functions as simple as possible. Simple has many advantages:

- Simple functions are likely to be human efficient: they will be easy
  to understand and to modify.
- Simple functions are likely to be computer efficient.
- Simple functions are less likely to be buggy, and bugs will be
  easier to fix.
- (Perhaps ironically) simple functions may be more general—thinking
  about the heart of the matter often broadens the application.

Functions can be

1. Correct.
2. An error occurs that is clearly identified.
3. An obscure error occurs.
4. An incorrect value is returned.

We like __category 1__. __Category 2__ is the right behavior if the inputs do not make sense, but not if the inputs are sensible. __Category 3__ is an unpleasant place for your users, and possibly for you if the users have access to you.  __Category 4__ is by far the worst place to be - the user has no reason to believe that anything is wrong. Steer clear of category 4.

### Your First Function

All R functions have three parts: 

* the `body()`, the code inside the function.

* the `formals()`, the list of arguments which controls how you can call the function.

* the `environment()`, the "map" of the location of the function's variables.

When you print a function in R, it shows you these three important components. If the environment isn't displayed, it means that the function was created in the global environment.

```{r}
myadd <- function(x, y) {
  cat(paste0("x = ", x, "\n"))
  cat(paste0("y = ", y, "\n"))
  x + y
}
myadd(1, 3)            # arguments by position
myadd(x = 1, y = 3)    # arguments by name
myadd(y = 3, x = 1)    # name order doesn't matter
```

- The body of the function is everything between the `{ }`.  Note this does the computation __AND__ returns the result.
- `x` and `y` are the arguments to the function.
- the environment this function lives in is the global environment. (We'll discuss environments more in the next section.)

```{block2, type='rmdtip'}
Even though it’s legal, I don’t recommend messing around with the order of the arguments too much, since it can lead to some confusion.
```

You can also specify default values for your arguments.  Default values _should_ be the values most often used.  `rnorm` uses the default of `mean = 0` and `sd = 1`.  We usually want to sample from the standard normal distribution, but we are not forced to. 

```{r}
myadd2 <- function(x = 3, y = 0){
  cat(paste0("x = ", x, "\n"))
  cat(paste0("y = ", y, "\n"))
  x + y
}
myadd2()              # use the defaults
myadd2(x = 1)
myadd2(y = 1)
myadd2(x = 1, y = 1)
```

By default the last line of the function is returned.  Thus, there is no reason to explicitly call `return`, unless you are returning from the function early.  Inside functions use `stop` to return error messages, `warning` to return warning messages, and `message` to print a message to the console.

```{r warning=TRUE, error=TRUE}
f <- function(age) {
  if (age < 0) {
    stop("age must be a positive number")
  }
  
  if (age < 18) {
    warning("Check your data.  We only care about adults.")
  }
  
  message(paste0("Your person is ", age, " years old"))
  invisible()
}

f(-10)
f(10)
f(30)
```

### Lazy Evaluation

R is lazy.  Arguments to functions are evaluated _lazily_, that is they are evaluated only as needed in the body of the function.

In this example, the function `f()` has two arguments: `a` and `b`.

```{r}
f <- function(a, b) {
  a^2
} 

f(2)     # this works
f(2, 1)  # this does too
```

This function never actually uses the argument `b`, so calling `f(2)` or `f(2, 1)` will not produce an error because the 2 gets positionally matched to a. It’s common to write a function that does not use an argument and not notice it simply because R never throws an error.

### The `...` Argument

There is a special argument in R known as the `...` argument, which indicate a variable number of arguments that are usually passed on to other functions. The `...` argument is often used when extending another function and you don’t want to copy the entire argument list of the original function

For example, a custom plotting function may want to make use of the default `plot()` function along with its entire argument list. The function below changes the default for the `type` argument to the value `type = "l"` (the original default was `type = "p"`).

```r
myplot <- function(x, y, type = "l", ...) {
        plot(x, y, type = type, ...)         ## Pass '...' to 'plot' function
}
```
The `...` argument is also necessary when the number of arguments passed to the function cannot be known in advance. This is clear in functions like `paste()` and `cat()`.

```{r}
args(paste)
args(cat)
```

Because both `paste()` and `cat()` print out text to the console by combining multiple character vectors together, it is impossible for those functions to know in advance how many character vectors will be passed to the function by the user. So the first argument to either function is `...`.

One catch with `...` is that any arguments that appear _after_ `...` on the argument list must be named explicitly and cannot be partially matched or matched positionally.

Take a look at the arguments to the `paste()` function.

```{r}
args(paste)
```

With the `paste()` function, the arguments `sep` and `collapse` must be named explicitly and in full if the default values are not going to be used.

## Environments & Scoping

An __environment__ is a collection of (symbol, value) pairs, i.e. `x` is a symbol and `3.14` might be its value. Every environment has a parent environment and it is possible for an environment to have multiple “children”. The only environment without a parent is the empty environment.  

__Scoping__ is the set of rules that govern how R looks up the value of a symbol. In the example below, scoping is the set of rules that R applies to go from the symbol `x` to its value `10`:

```{r}
x <- 10
x
```

R has two types of scoping: lexical scoping, implemented automatically at the language level, and dynamic scoping, used in select functions to save typing during interactive analysis. We discuss lexical scoping here because it is intimately tied to function creation. Dynamic scoping is an advanced topic and is discussed in [Advanced R](http://adv-r.had.co.nz).

How do we associate a value to a free variable? There is a search process that occurs that goes as follows:

If the value of a symbol is not found in the environment in which a function was defined, then the search is continued in the parent environment.
The search continues up the sequence of parent environments until we hit the top-level environment; this usually the global environment (workspace) or the namespace of a package.
After the top-level environment, the search continues down the search list until we hit the empty environment.
If a value for a given symbol cannot be found once the empty environment is arrived at, then an error is thrown.

```{r}
x <- 0
f <- function(x = -1) {
  x <- 1
  y <- 2
  c(x, y)
}

g <- function(x = -1) {
  y <- 1
  c(x, y)
}

h <- function() {
  y <- 1
  c(x, y)
}
```

What do the following return?

- `f()`
- `g()`
- `h()`
- `g(h())`
- `f(g())`
- `g(f())`

Unlike most languages you can define a function within a function.  This nested function only lives inside the parent function.

```{r error=TRUE}
make.power <- function(n) {
  pow <- function(x) {
    x^n 
  }
  pow
}

make.power(4)
cube <- make.power(3)
square <- make.power(2)

x <- 1
n <- 2
pow(x=4)
```

## Exercises

1. Browse this [vocabulary list](http://adv-r.had.co.nz/Vocabulary.html) and read the help file for functions that interest you.
2. Re-run the three cases in the For loop section with `x <- NULL`
3. Vectorization / function practice.

We'll calculate pi using the Gregory-Leibniz series. Mathematicians will be quick to point out that this is a poor way to calculate pi, since the series converges very slowly. But our goal is not calculating pi, our goal is examining the performance benefit that be be achieved using vectorization.

Here is a formula for the Gregory-Leibniz series:

\begin{equation}
1 - \frac{1}{3} + \frac{1}{5} - \frac{1}{7} + \frac{1}{9} - \frac{1}{11} + \cdots = \frac{\pi}{4}
\end{equation}

Here is the Gregory-Leibniz series in summation notation:

\begin{equation}
\sum_{\text{n}=0}^{\infty} \frac{(-1)^n}{2\cdot n + 1} = \frac{\pi}{4}
\end{equation}

The straightforward implementation using an R loop would look like this:

```{r}
GL_naive <- function (limit) {
  p = 0
  for (n in 0:limit) {
    p = (-1)^n/(2 * n + 1) + p
    }

  4*p
}

N <- 1e7
system.time(pi_est <- GL_naive(N))
pi_est
```

Your task is to vectorize this function.  Do not use any looping or apply functions.  This one is a bit tricky.  Hint: It may be easier to think about it in terms of the series notation and not the summation notation.

```{r eval=FALSE}
GL_vect <- function(limit) {
  # your code here
  # use only base functions and no looping mechanisms
}
```
